---
title: "United Kingdom’s Carbon Monoxide emissions Project"
format: html
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

***Question :- “What trends and turning points can be identified in United Kingdom's Carbon Monoxide emissions over time?”***

```{r}
library(readr)
library(tidyverse)
air_pollutants <- read_csv("https://ourworldindata.org/grapher/long-run-air-pollution.csv?v=1&csvType=full&useColumnShortNames=true")
head( air_pollutants)
```
- Dataset now has data about various countries and regions.  

```{r}
# Selecting only CO emission column and UK for rows for the analysis. 
air_pollutants_UKCO <- air_pollutants %>% 
  select(entity, code, year, emissions__pollutant_co__sector_all_sectors) %>% 
  rename("CO" = "emissions__pollutant_co__sector_all_sectors") %>% 
  filter( entity == "United Kingdom")

#Cleaning Dataset
colSums((is.na(air_pollutants_UKCO))) #No missing data found

air_pollutants_UKCO <- air_pollutants_UKCO %>% #Removing duplicate rows
  distinct()
```
- Dataset now is filtered for United Kingdom and free of duplicates.  
```{r}
air_pollutants_UKCO <- air_pollutants_UKCO %>% 
  mutate(Colour = case_when( year < 1921 ~ "Pre-war",
                             year >= 1921 & year < 1956 ~ "Post-war",
                             year >= 1956 & year < 1990 ~ "Clean Air Act",
                             year >= 1990 ~ "UK Act"
  ))


ggplot() +
  geom_line( data = air_pollutants_UKCO, 
    aes( x = year, y = CO, colour = Colour),
    lwd = 0.8
    ) +
  labs(
    x = "Year",
    y = "Carbon Monoxide Emissions per year",
    title  = "Carbon Monoxide Emissions in United Kingdom per year in tonnes",
    subtitle = "From 1750 - 2022",
    caption = "Source : Hoesly et al - Community Emissions Data System (CEDS)"
  ) + 
  geom_point(
    data = air_pollutants_UKCO%>% filter(year %in% c(1921, 1956, 1990)),
    aes(x = year, y = CO, color = Colour),
    size = 3
  ) +
annotate(
"text",
x = 1921, 
y = 4500000, 
label = "Post war\nIndustrial expansion\n(1921)",
hjust = 0,
size = 3.5,
colour = "red"
) + 
  annotate(
"text",
x = 1965, 
y = 14000000, 
label = "Introduction of\nClean Air Act (1956)",
hjust = 0,
size = 3.5,
colour = "goldenrod"
) + 
  annotate(
"text",
x = 1975, 
y = 10500000, 
label = "UK\nenvironmental\nprotection act\n(1990)",
hjust = 0,
size = 3.5,
colour = "forestgreen"
) + 
  scale_colour_manual(
  values = c("Pre-war" = "steelblue", "Post-war" = "red", "Clean Air Act" = "goldenrod", "UK Act" = "forestgreen")
  ) +
scale_x_continuous(breaks = seq(1750, 2022, by = 20),
  minor_breaks = seq(1750, 2022, by = 5)) +
  expand_limits(y=0) +
  theme_minimal() +
   theme(
   plot.caption = element_text(size=6,color="gray",face="italic")
  )
```
- Graph shows carbon monoxide emissions from 1750 to 2023. It clearly shows there was a serious increase in emissions and it reached aa peak at some year and then it started to reduce.  
- Further, the graph shows different region of time where emissions has significant change. Red section are shows how after 1921's post-war industrial expansion emission started to increase rapidly.    
- And it reached to its peak until Clean Air Act was introduced in 1956 and emissions decreased after that. Which is highlighted by the yellow region.  
- Then, 1990's United Kingdom Environmental Protection Act was the final nail, from where it showed a further decrease. Which is highlighted by the green region.  
```{r}

ggplot() +
   geom_vline(xintercept = 1921, linetype = "dashed", colour = "grey50") +
   geom_vline(xintercept = 1956, linetype = "dashed", colour = "grey50") +
   geom_vline(xintercept = 1990, linetype = "dashed", colour = "grey50") +
  geom_line( data = air_pollutants_UKCO,
    aes( x = year, y = CO), colour = "steelblue",
    lwd = 0.8)  +
geom_smooth(
  data = air_pollutants_UKCO,
  aes(x = year, y = CO),
  method = "gam",
  colour = "red",
  linewidth = 1,
  se = FALSE
) + 
  labs(
    x = "Year",
    y = "Emission of Carbon Monoxide in UK over the years",
    title  = "Emissions of Carbon Monoxide in United Kingdom per year in tonnes",
    subtitle = "From 1750 - 2022",
    caption = "Source : Hoesly et al - Community Emissions Data System (CEDS)\n1921* - Post was Industrial Expansion\n1956* - Introduction to clean air act\n1990* - UK environmental protection act"
  ) +
  annotate(
"text",
x = 1897, 
y = 14000000, 
label = "1921*",
hjust = 0,
size = 3.5,
colour = "grey50"
) + 
  annotate(
"text",
x = 1932, 
y = 14000000, 
label = "1956*",
hjust = 0,
size = 3.5,
colour = "grey50"
) + 
  annotate(
"text",
x = 1966, 
y = 14000000, 
label = "1990*",
hjust = 0,
size = 3.5,
colour = "grey50"
) +
    scale_x_continuous(breaks = seq(1750, 2022, by = 20),
  minor_breaks = seq(1750, 2022, by = 5)) +
  scale_y_continuous(breaks = seq(0, 20000000, by = 1000000)) +
  expand_limits(y=0) +
 theme_minimal() +
  theme(
   plot.caption = element_text(size=8,color="gray",face="italic")
  )

```
- Smoothened graph clearly shows that the period between 1921 and 1956 had the most emissions. And after Clean Air Act and United Kingdom Environmental Protection Act there was a drastic decrease in emissions.  
- It alos suggests that emssions trend is a curve and not a line.  
```{r}
model_co_poly <- lm(CO ~ year + I(year^2), data = air_pollutants_UKCO)
summary(model_co_poly)

b1 <- coef(model_co_poly)["year"]
b2 <- coef(model_co_poly)["I(year^2)"]

turning_point <- -b1 / (2 * b2)
turning_point

ggplot(air_pollutants_UKCO,
       aes(x = year, y = CO)
      ) +
  geom_point(color = "darkgrey") +
  labs(
    x = "Year",
    y = "CO Emissions",
    title = "CO Emissions vs Year (UK)",
  ) +
  geom_smooth(method = lm, 
              formula = y ~ poly(x,2), 
              color = "red", se = FALSE
              ) +
  theme_minimal()
```
- The positive coefficient for year and negative coefficient for year squared indicate an there was an increase at start and then followed by a decline after reaching a peak.  
- The model explains approximately 63% of the variation in CO emissions (R² = 0.63), indicating a reasonably strong relationship between time and emissions.  
- Model indicated year 1925.342 was the year of maximum emissions. This was the turning point from where emissions started to decrease.   


***Conclusion***  

- This analysis examined long-run carbon monoxide emissions in the United Kingdom.  
- The visualisations show that emissions increased over time before reaching a peak and then declining in later years.  
- A quadratic regression model supported this non-linear pattern, indicating that CO emissions did not follow a simple linear trend.  
- Overall, the results suggest a strong historical shift in UK emissions over time, consistent with industrial growth followed by later environmental control and technological change.
